---
title: |
  Data Visualization and Parquet Files
editor: source
categories: ["Class Materials"]
format: revealjs
date: 2026-10-01
---

# Today's Goals {visibility="hidden"}

```{r setup}
#| include: false
# Load local .env from project root regardless of folder depth

here::i_am(".env")
env_path <- here::here(".env")
if (file.exists(env_path)) {
  readRenviron(env_path)
} else {
  stop(paste("Unable to locate .env file at ", env_path))
}

# Extract Google Drive path  
course_path <- Sys.getenv("GDRIVE_PATH")
if(!dir.exists(course_path)) {
  stop(paste("Google drive folder ", course_path, " does not exist. Check the .env file."))
}

data_path = file.path(course_path,"Data")
library(tidyverse)
library(arrow)
```

## Today's Goals {.smaller}

The **Parquet** data file format

- Understand the advantages and disadvantages of the Parquet file format
- Reading and writing Parquet files in R

Additional Data Visualization using `ggplot()`

- Construct data visualization using `ggplot()`
- Creating legends
- Save visualizations for use in other documents

We will continue to use the FAO data in

`AGNR-AREC280_Fall2026/Data/FAO`

# Modern Data File Formats

## CSV has been a *de facto* standard for data {.smaller}

**The Bottleneck:**

- Modern datasets like high-resolution climate rasters or satellite indices overwhelm plain text formats like CSV
  - e.g., The FAO data is nearly 400 MB in CSV format
- Data stored in purely text format
  - Requires software "guess" properties of the data
  - Numeric data needs to be translated from text
  - Repeated text values written out in full, each time
- CSV files are
  - Slow to read
  - Large
  - Require guesses or manual intervention to read correctly

## Modern Big Data Storage

**The Next Generation:**

Define a data file format for modern big data

- **Portable**: Works across data processing systems
- **Fast**: Information you need loads quickly
- **Compact**: Data stored efficiently for small files
- **Consistent**: All software interprets the data identically without guessing

## The Parquet Format

[**Parquet**](https://en.wikipedia.org/wiki/Apache_Parquet) (PAR-kay) is a modern data file format for storing tidy data frames

- Created in 2013
- Specification is free and open source
- Managed by the [Apache Foundation](https://en.wikipedia.org/wiki/The_Apache_Software_Foundation)
- Implemented in most data analysis software

## Parquet Advantages: Column-orientation

**Column-oriented storage**

- In CSV files data are oriented in rows
  - You must read all of one row before moving on to the next
  - Inefficient if you only want data from specific columns
- Parquet files use a column format
  - Only need to read data for the columns you request

## Parquet Advantages: Automatic Compression

**Automatic Compression**

- A CSV must write out every value in every cell as text
- Parquet automatically compresses information
  - Column-oriented files are more compressible in general
  - [Dictionary compression](https://en.wikipedia.org/wiki/Dictionary_coder) of frequent values
  - [Run-length encoding](https://en.wikipedia.org/wiki/Run-length_encoding) for repeated values
  - [Delta-encoding](https://en.wikipedia.org/wiki/Delta_encoding) efficiently packs sequential numbers/timestamps

## Parquet Advantages: Strict Schema

**Strict Schema**

- A CSV file doesn't contain any information about data types
- Parquet files explicitly define data types of each column
  - Integer, Floating point, Character, Boolean, Date, and Timestamp data types
- Parquet uses explicit formats for each data type (e.g., [IEEE-754 floating point](https://en.wikipedia.org/wiki/IEEE_754))
- Any data processing package interprets each value the exact same way
- Explicit support for missing (`NA`) values

## Parquet Disadvantages

- Binary files aren't human-readable
- Column-oriented format makes "appending" data slow
  - Poorly suited for cases where you frequently update data
- Performance gains only materialize on larger files
  - 1000s of observations
  - 10s or 100s of columns

## Parquet in this Class

In this class I will provide Parquet versions of most data files

- Data files contain many, repeated text values
- Very large CSV files
- Parquet much smaller, so they load quickly
- Strict schema, so everyone gets the same data
- ELMS and gmail understand they are data files

# Loading Parquet Files in R

## Loading Parquet files in R

You can load parquet files in R

- Use `read_parquet()`
- Requires the `arrow` library

Load `FAO/parquet/FAO_crop_livestock.parquet` into R.

::: callout-tip
### New Package

You will need to install the `arrow` package if you haven't already
:::

## Load the FAO Parquet file

```{r load_faostat}
#| echo: true
library(arrow)
read_parquet(
  file.path(data_path,"FAO","parquet","FAO_crop_livestock.parquet")
) -> faostat
```

This is the full FAO data file

- Full time series (1961 - 2025)
- Full set of countires
- Full set of commodities

Let's start by focusing on a single measurement of one crop in one country

# Wrangling the FAOSTAT Data

## Choose your adventure: Country {.smaller}

Create a pipeline to construct a single crop data frame

**Step 1**

- Use `distinct()` to look at countries (`Area`)
- Choose one you are interested in
  - Write a `filter()` command that would limit the data to that country

::: callout-tip
### Displaying long results

R limits results in the console to 10 - 20 lines.

Pipe your result to `print(n=Inf)` to display all rows
:::

## My Adventure: Country

I chose to look at France

```{r filter_country}
#| echo: true
faostat |>
  filter(Area == "France") 
```

## Choose your adventure

Create a pipeline to construct a single crop dataset

**Step 2**

- Pipe your previous result to `distinct()` to look at commodities (`Item`)
- Choose one commodity you are interested in
  - Write a `filter()` command that would limit the data to that commodity

## My Adventure: Commodity

I chose to look at Wine in France

```{r filter_commodity}
#| echo: true
faostat |>
  filter(Area == "France") |>
  filter(Item == "Wine") 
```

## Choose your adventure

Create a pipeline to construct a single crop dataset

**Step 3**

- Pipe your previous result to `distinct()` to look at measurements (`Element`)
- Choose one measurement you are interested in
  - Write a `filter()` command that would limit the data to that measurement

## My Adventure: Measurement

I chose to look at Wine in France

```{r filter_measurement}
#| echo: true
faostat |>
  filter(Area == "France") |>
  filter(Item == "Wine") |> 
  filter(Element == "Production")
```

## Choose your adventure

Create a pipeline to construct a single crop dataset

**Step 4**

- Save this filtered data frame to a new tibble

```{r save_filtered}
#| echo: true
faostat |>
  filter(Area == "France") |>
  filter(Item == "Wine") |> 
  filter(Element == "Production") -> fao_filtered
```

## Check your work

Examine you filtered data frame to be sure you have information you expect

- `glimpse()`
- `summary()`
- `distinct()` or `count()`

## Visualize {.smaller}

Use your filtered data frame to create a visualization

- Year on the horizontal axis
- Your measurement on the vertical axis
- Black dots for the value of measurement in each year (using `geom_point`)
- A blue line showing the trend (using `geom_smooth`)
- Clear labels throughout the visualization

::: callout-tip
### Removing error band from `geom_smooth`

You can remove the error band from `geom_smooth` layers by setting the layer property `se=FALSE`
:::

## My visualization

```{r viz_1}
#| echo: true
#| code-fold: true
ggplot(
  data=fao_filtered,
  mapping=aes(x=Year,y=Value)
) +
  geom_line(color="black") +
  geom_smooth(color="blue",se=FALSE) +
  labs(
    title="Wine Production in France",
    x="Year",
    y="Production (mt)",
    caption="Source: United Nations FAOSTAT"
  )
```

# On-the-fly Calculations in Visualizations

## A tweak to the labels

- The values on my vertical axis are in scientific notation
  - R defaults to using it when values are larger than 1000
  - Or smaller than 0.1
- Sometimes difficult to read
- Options for improving:
  - Force R to format numbers in fixed-comma format (8,000,000)
  - Rescale the axis so numbers are smaller (**million mt** vs **mt**)

## Rescale values on-the-fly

The cleanest choice is to rescale the vertical axis

- Show "million mt" instead of "mt"
- Requires dividing `Value` by 1,000,000
- In `ggplot` you can do this math "on-the-fly" in a visualization

```{r}
#| eval: false
#| echo: true
...
  aes(x=Year,y=Value/1000000)
...
```

- Any mathmatical operation is supported
- Summary functions (e.g., `mean()`) will compute the value over all data in that layer

## My visualization

```{r viz_1_rescale}
#| echo: true
#| code-fold: true
ggplot(
  data=fao_filtered,
  mapping=aes(x=Year,y=Value/1000000)
) +
  geom_line(color="black") +
  geom_smooth(color="blue",se=FALSE) +
  labs(
    title="Wine Production in France",
    x="Year",
    y="Production (million mt)",
    caption="Source: United Nations FAOSTAT"
  )
```

# Customizing an Axis

## Visualization Scales

- R automatically adjusts the scale of each axis to fit the data
  - Minimum value slightly smaller than smallest value
  - Maximum value slightly larger than largest value
- Can exaggerate how much values change over time
- In the previous visualization, you may want to extend the axis to zero

## Adjusting Visualization Axes

- Adjust axis properties using **scale** functions
- Take the form: `scale_<axis>_<type>()`
- `<axis>` is `x` or `y`
- `<type>` is the axis type
  - Determines how the axis will be drawn
  - Depends on the type of data you're plotting

## Adjusting Visualization Axes

::: callout-tip
### Axis Scale Resources

The **Help** feature in RStudio has excellent documentation, start by typing "scale\_" in the search box.

There is also an excellent resource for using scales at

<https://ggplot2-book.org/scales-guides.html>
:::

## Adjusting a Continuous Axis

The vertical axis in our visualization is **continuous**

- Customize using `scale_y_continuous()`
- Read the **help** for this function
- There will be some things that make sense and some thing less so
- Important options (look closely at them)
  - `limits`
  - `breaks`
  - `minor_breaks`

## Adjusting a Continuous Axis

Arguments for `scale_y_continuous()`

- `limits`
  - A 2-element vector with the starting and ending value of the axis
- `breaks`
  - A vector with the major (labeled) break values on the axis
- `minor_breaks`
  - A vector with the minor (unlabeled) break values on the axis

## Adjusting a Continuous Axis

I am going to

- Set the axis to start at 0 and go to 10 (`limits`)
- Have major breaks at even numbers (`breaks`)
- Have minor breaks at odd numbers (`minor_breaks`)

::: callout-tip
### Creating Vectors

Create vectors with the `c()` function

You can create a sequence of numbers in a vector with the `seq()` function
:::

Make changes you think are appropriate for your visualziation!

## Visualization with Updated Axis

```{r example_viz_2}
#| echo: true
#| code-fold: true
ggplot(
  data=fao_filtered,
  mapping=aes(x=Year,y=Value/1000000)
) +
  geom_line(color="black") +
  geom_smooth(color="blue",se=FALSE) +
  labs(
    title="Wine Production in France",
    x="Year",
    y="Production (million mt)",
    caption="Source: United Nations FAOSTAT"
  ) +
  scale_y_continuous(
    limits=c(0,10),
    breaks=seq(0,10,2),
    minor_breaks=seq(0,10,1)
  )
```

# Storing Visualizations

## Storing visualizations

You can store a `ggplot` visualization using an assignment operator (e.g., `->`)

- Stores everything needed to create the visualization in a named variable
- Does not render the visualization
- You can render/display it later with `print()`
- Why would you do this?
  - Create many visualizations then print/save them all at once
  - You can modify a stored visualization by adding more layers or attributes

## Storing a Visualization

This stores my visualization, but does not show it!

```{r define_viz_2}
#| echo: true
ggplot(
  data=fao_filtered,
  mapping=aes(x=Year,y=Value/1000000)
) +
  geom_line(color="black") +
  geom_smooth(color="blue",se=FALSE) +
  labs(
    title="Wine Production in France",
    x="Year",
    y="Production (million mt)",
    caption="Source: United Nations FAOSTAT"
  ) +
  scale_y_continuous(
    limits=c(0,10),
    breaks=seq(0,10,2),
    minor_breaks=seq(0,10,1)
  ) -> production_viz
```

# `ggplot` Themes

## Using Themes

Overall appearance of a visualization is controlled by **themes**

- Themes are customizable
- R has some built-in themes that combine common options
- Each is named `theme_<name>()`
- Add to a visualization to apply
- Default theme is `theme_grey()`

## Example Themes `theme_bw()`

```{r viz_2_theme_bw}
#| echo: true
#| code-fold: true
production_viz +
  theme_bw()
```

## Example Themes `theme_light()`

```{r viz_2_theme_light}
#| echo: true
#| code-fold: true
production_viz +
  theme_light()
```

## Example Themes `theme_dark()`

```{r viz_2_theme_dark}
#| echo: true
#| code-fold: true
production_viz +
  theme_dark()
```

## Example Themes `theme_minimal()`

```{r viz_2_theme_minimal}
#| echo: true
#| code-fold: true
production_viz +
  theme_minimal()
```

# Saving Visualizations

## Saving Visualizations

Visualizations are most useful if you can share them

- R Studio's Plots viewer allows saving visualizations
  - Several steps
  - Not Automated
  - Required every time you update the visualization
- You can add code to save a visualization with `ggsave()`
  - Visualization will be saved each time you run your R code
  - Apply specific options to the saved file (file format, size)

## Using `ggsave`

To save a visualization use

`ggsave(<output_path>)`

- R determines the output format using the file extension
  - e.g., `.png`, `.jpg`, `.pdf`
- Some important optional arguments are:
  - `plot`: A variable containing the plot to save
    - Otherwise, the default is the last plot generated
  - `width`: Width of the saved image in inches
  - `height`: Height of the saved image in inches

## Example Saving a Visualization

```{r}
#| echo: true
#| eval: false

# Set a path to a directory where you can save files
homedir_path = "/home/username/Documents/AREC280"

ggsave(
  file.path(homedir_path,"wine_production.png"),
  plot=production_viz,
  width=16,
  height=9
)
```

::: callout-warning
### No write permissions to the Data folder

You do not have permission to write files to the Data folder in the class Google drive. Attempting to save a file to that folder will result in a **Permission Denied** error.

Be sure to save to a folder on your local computer.
:::

```{r}
#| echo: false
#| eval: true

# Save some sample files to show
ggsave(
  "wine_production.png",
  plot=production_viz,
  width=16,
  height=9
)

ggsave(
  "wine_production.pdf",
  plot=production_viz,
  width=16,
  height=9
)
```

## Choosing a Visualization File Format {.smaller}

You have several choices for file formats, which have advantages and disadvantages

- [**PDF**](https://en.wikipedia.org/wiki/PDF) files best for publication quality graphics
  - Are [vector images](https://en.wikipedia.org/wiki/Vector_graphics), they scale up flawlessly to very large image sizes
  - Complicated visualizations make large output file that load slowly
- [**PNG**](https://en.wikipedia.org/wiki/PNG) files best for general use
  - Are [raster images](https://en.wikipedia.org/wiki/Raster_graphics), they will get blocky as you scale them up
  - Files will be a fixed size no matter how complicated the visualization
- [**JPG**](https://en.wikipedia.org/wiki/JPEG) files best for websites/text messages
  - Also raster images but use [Lossy Compression](https://en.wikipedia.org/wiki/Lossy_compression)
  - Files tend to be very small, nearly anyone can view them
  - Underlying compression algorithm can lose details
- There are other, more niche formats available too

## Next Class

- More Data Visualization
