---
title: |
  Data Visualization with Multiple Groups
editor: source
categories: ["Class Materials"]
format: revealjs
order: 3
---

# Today's Goals {visibility="hidden"}

```{r setup}
#| include: false
# Load local .env from project root regardless of folder depth

here::i_am(".env")
env_path <- here::here(".env")
if (file.exists(env_path)) {
  readRenviron(env_path)
} else {
  stop(paste("Unable to locate .env file at ", env_path))
}

# Extract Google Drive path  
course_path <- Sys.getenv("GDRIVE_PATH")
if(!dir.exists(course_path)) {
  stop(paste("Google drive folder ", course_path, " does not exist. Check the .env file."))
}

data_path = file.path(course_path,"Data")
library(tidyverse)
library(arrow)
```

## Today's Goals

Today we will

- Construct a subset of the FAO data with multiple countries
- Construct visualizations that compare value across countries
- Understand options for customizing legends and color schemes in ggplot

We will continue to use the FAO data in

`AGNR-AREC280_Fall2026/Data/FAO`





# Construct an FAO Subset with multiple countries

## Expanding our FAO subset

Today we're going to create visualizations that compare data across different 
groups using the FAO data

- Previously we'd created a subset of the data for a single country-product
- Today we'll need multiple groups to compare
- We will construct a subset of the FAO data 
  - A single product
  - A single measurement
  - Multiple countries

## Create a subset of the FAO data {.smaller}

Create a subset of the FAO data limited to

- **Wheat** as the commodity
- **Production** as the measurement
- For the following countries:
  - **Sweden**
  - **Norway**
  - **Finland**
  
Save the subset to a variable named `fao_wheat`. 

After you have created it, use data summary commands to check whether your 
new dataset contains what you expect. 
  
## Loading the FAO Data

First load the FAO data from the Parquet file

```{r load_fao}
#| echo: true
#| code-fold: true
# Load required libraries
library(tidyverse)
library(arrow)

# Load the full fao dataset
read_parquet(
  file.path(data_path,"FAO","parquet","FAO_crop_livestock.parquet")
) -> faostat

# glimpse to be sure we've loaded the data
glimpse(faostat)
```

## Creating the `fao_wheat` subset

```{r create_subset}
#| echo: true
#| code-fold: true
# Create the subset
faostat |>
  # Limit to Wheat as the commodity
  filter(Item == "Wheat") |>
  # Limit to Production as the measurement
  filter(Element == "Production") |>
  # Limit to Sweden, Norway, and Finland
  filter(Area %in% c("Sweden","Norway","Finland")) -> fao_wheat

# glimpse a few rows
glimpse(fao_wheat)
```

## Check our work

- There are many things you could do to check you've done what you expected
- Here I'll count observations by country

```{r check_countries}
#| echo: true
#| code-fold: true
fao_wheat |>
  count(Area)
```

What other checks could you perform?


# Visualizations over groups

## A first visualization

Create a visualization like in previous classes

- Year on the horizontal axis
- Production on the vertical axis

Does it make sense, what has gone wrong?


## Visualization over groups - Bad

```{r viz_1}
#| echo: true
#| code-fold: true
ggplot(
  data=fao_wheat,
  mapping=aes(x=Year,y=Value/1000000)
) +
  geom_line(color="black",linewidth=1) +
  labs(
    title="Nordic Wheat Production",
    x="Year",
    y="Production (million mt)",
    caption="Source: United Nations FAOSTAT"
  ) +
  theme_bw()
```

## Visualization over groups - What went wrong

- Our new data has three observations for each year
- One for each country
- Plots each point but doesn't differentiate by country

## Visualization over groups - Set aesthetics

You can tell ggplot how to group data using aesthetics

- The `group=<var>` aesthetic will create a new layer for each unique value of `<var>`
- The `color=<var>` aesthetic will use a different color for observations depending on the value of `<var>`

Modify the previous visualization to add a `color=Area` aesthetic

## Visualization over groups - `color` aesthetic

```{r viz_1_color}
#| echo: true
#| code-fold: true
ggplot(
  data=fao_wheat,
  mapping=aes(x=Year,y=Value/1000000,color=Area)
) +
  geom_line(linewidth=1) +
  labs(
    title="Nordic Wheat Production",
    x="Year",
    y="Production (million mt)",
    caption="Source: United Nations FAOSTAT"
  ) +
  theme_bw() -> viz_wheat_color

print(viz_wheat_color)
```


## Recap - `color` aesthetic

Adding the `color=Area` aesthetic

- Creates separate lines for each value of `Area`
- Assigns a different color for each value of `Area`
- Creates a legend showing the colors corresponding to each value of `Area`

Are these good colors to use?


# Setting Custom Colors

## Setting custom colors

- R uses a default color palate for colors
- It is not ideal
- You can override default colors with 
  - `scale_color_manual()`: You define the colors
  - `scale_color_brewer()`: Use color scales from `RColorBrewer`
  - `scale_color_viridis_<b|c|d>()`: Use color scales from `viridis`
  
## Setting custom colors - Brewer

By default Brewer uses blue colors and intensity
```{r viz_1_color_brewer}
#| echo: true
#| code-fold: true
viz_wheat_color + 
  scale_color_brewer()
```

## Brewer palates

The RColorBrewer is optimized for maps and has a variety of palates set using `palette="<palate>"`

- Sequential (e.g., `YlOrRd`, `PuRd`)
- Qualitative (e.g., `Set3`, `Dark2`)
- Diverging (e.g., `Spectral`, `PuOr`)

A gallery is available in the [R Graph Gallery](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html)

## Setting custom colors - Brewer Set3

`Set3` is a categorical scale
```{r viz_1_color_brewer_set3}
#| echo: true
#| code-fold: true
viz_wheat_color + 
  scale_color_brewer(palette="Set3")
```


## Setting custom colors - Viridis

Here is Viridis D

```{r viz_1_color_viridis_d}
#| echo: true
#| code-fold: true
viz_wheat_color + 
  scale_color_viridis_d()
```
## Viridis Palettes

Viridis palates are designed 

- To be clear even if the viewer has common forms of colorblindness
- To be clear when converted to greyscale
- To have strongly contrasting colors with a logical order of progression

## Viridis Scales {.smaller}

The correct Viridis scale to use depends on your data:

- `scale_color_viridis_b()`
  - **Binned**: Discrete *with a defined order/ranking*
  - Adjacent values have similar colors
- `scale_color_viridis_c()` 
  - **Continuous**: A range where any value is possible
- `scale_color_viridis_d()` 
  - **Discrete**: Discrete *but unordered*
  - Maximizes contrast between colors

What type of data do we have here?

## Viridis Colors

Viridis has several color range options

- Set using `option="<name>"`
- Possible names include `magma`, `inferno`, `viridis`, `rocket`, etc.

See all the options in the [Viridis Documentation](https://search.r-project.org/CRAN/refmans/viridisLite/html/viridis.html)

## Other Important Viridis Options

The Viridis scale functions also have some other important options:

- `direction=[-1|1]`
  - Sets the order of the colors. `-1` reverses them
- `na.value="<color>"`
  - Sets the color used when a value is `NA`
  - Set to `NA` for transparent


## Setting Custom Colors - Manual

You may occasionally want to set manual colors

- Use `scale_color_manual()`
- Set the `value=` argument to a **named vector**
- Each element of the vector is a `"<value>"="<color>"` pair
- e.g., `"Sweden"="yellow"`


## Setting Customs Colors - Manual, easy

Set colors for each country to a color like one in their flag.

```{r viz_1_color_manual_easy}
#| echo: true
#| code-fold: true
viz_wheat_color + 
  scale_color_manual(
    values=c(
      # Each pair is a country name (from the data)
      # And the the name of color in their flag
      "Finland" = "blue", #Finland blue
      "Sweden" = "yellow", #Sweden Yellow
      "Norway" = "red" #Norway Red
    )
  )
```

## Setting Customs Colors - Manual, hard

Set colors for each country to a color in their flag using [web colors](https://en.wikipedia.org/wiki/Web_colors).

```{r viz_1_color_manual_hard}
#| echo: true
#| code-fold: true
viz_wheat_color + 
  scale_color_manual(
    values=c(
      # Each pair is a country name (from the data)
      # And the hex code for a color in their flag
      "Finland" = "#002F6C", #Finland blue
      "Sweden" = "#FECB00", #Sweden Yellow
      "Norway" = "#BA0C2F" #Norway Red
    )
  )
```

## Best practices for setting colors

- R's default color choices are not ideal, particularly if more than a few colors
- Both RColorBrewer and Viridis were designed by professionals
- Use their palettes when possible
- Only manually define colors if a specific color is important

# Adjusting the legend appearance

## Legend Defaults

By default, ggplot

- Titles the legend after the corresponding aesthetic (e.g., "color")
- Places it on the right in vertical format

These defaults are not always optimal

## Changing the legend title

- You can change the legend title in `labs()`
- The label name is the name of the corresponding aesthetic (e.g., "color")
- e.g., `labs(color="Country")`

Change the legend title to "Country" in your visualization


## Updated Legend name
```{r update_legend_title}
#| echo: true
#| code-fold: true
viz_wheat_color +
  scale_color_viridis_d() +
  labs(color="Country")
```

## Starting from scratch
```{r add_legend_title}
#| echo: true
#| code-fold: true
ggplot(
  data=fao_wheat,
  mapping=aes(x=Year,y=Value/1000000,color=Area)
) +
  geom_line(linewidth=1) +
  labs(
    title="Nordic Wheat Production",
    x="Year",
    y="Production (million mt)",
    caption="Source: United Nations FAOSTAT",
    color="Country"
  ) +
  scale_color_viridis_d() +
  theme_bw() -> viz_wheat_color

print(viz_wheat_color)
```

# Changing the Legend Position

## Adjusting Legend Attributes

Many attributes of the way a legend is displayed are controlled by the plot theme

- Change individual settings of the theme with `theme()`
- Theme takes a series of named argument-value pairs
- Named arguments define attributes of the theme
- More complicated options controlled by `guides()`

See [Tidyverse ggplot2 3.5.0: Legends](https://tidyverse.org/blog/2024/02/ggplot2-3-5-0-legends/)
for many examples

## Changing the legend position

The legend placement is controlled by theme attribute `legend.position=`

- Possible values `"left"`, `"right"`, `"bottom"`, `"top"`
- `"none"` turns off the legend
- `"inside"` requires `legend.position.inside=c(x,y)`
  - Places the legend inside at the specified `x` and `y` coordinates
  
Move the legend to the bottom of your visualization.

- How does it change?


## Move the legend to the bottom

```{r legend_bottom}
#| echo: true
#| code-fold: true
viz_wheat_color +
  theme(
    legend.position = "bottom"
  )
```


# Using groups for calculations

## Group-level calculations

- Recall that you can place mathematical expressions in aesthetics
- R will evauluate them and use the result to define the data displayed
- You can use summary functions like `mean()`
- By default R will compute the summary over the full dataset
- Add the `group=` aesthetic to perform the calculation over each group

## Group-level Calculations

There may be occasions where you would like to perform a calculation
specific to each group and visualize that result

- A few options
  - Use `mutate()` combined with `if_else()` or `case_when()`
  - Use `group_by()` and summary functions
  - Use `group=` along with calculations in an aesthetic

Let's think about implementing the `group=` method

## Example of a Group-Level Calculation

What will the following code do?
```{r}
#| eval: false
#| echo: true

ggplot(data=fao_wheat) +
  geom_line(mapping=aes(
    x=Year,
    y=(Value - mean(Value))/sd(Value),
    group=Area,
    color=Area
  ))
```

## Group-level Calculation in a Visualization

```{r viz_z_score}
#| code-fold: true
#| echo: true

ggplot(data=fao_wheat) +
  geom_line(mapping=aes(
    x=Year,
    y=(Value - mean(Value))/sd(Value),
    group=Area,
    color=Area
  )) +
  theme_bw() + 
  theme(legend.position = "bottom") +
  labs(
    title="Nordic Wheat Production",
    x="Year",
    y="Relative Production (z-score)",
    caption="Source: United Nations FAOSTAT",
    color="Country"
  ) +
  scale_color_manual(
    values=c(
      "Finland" = "#002F6C", #Finland blue
      "Sweden" = "#FECB00", #Sweden Yellow
      "Norway" = "#BA0C2F" #Norway Red
    )
  )
  
```

# Wrap up {visibility="hidden"}

## Next Class

- **Data Transformations:** Reshaping Data
